---
title: "Nested convolution reparameterisation"
author: "Adam Howes"
---

```{r}
library(rstan, quietly = TRUE)
library(INLA, quietly = TRUE)
library(ggplot2, quietly = TRUE)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

Grouping structure can be either cross-classified or nested.
Here are some [slides](http://www.utstat.toronto.edu/~brunner/oldclass/appliedf12/lectures/2101f12RandomNested.pdf) from UT about it.

In this example we follow the nested model [eggs example](https://people.bath.ac.uk/jjf23/mixchange/nested.html) of Julian Faraway, who provides code to fit the model in (amongst other things) [`R-INLA`](https://people.bath.ac.uk/jjf23/inla/nested.html) and [`rstan`](https://people.bath.ac.uk/jjf23/stan/nested.html).

```{r}
data(eggs, package = "faraway")
summary(eggs)
head(eggs)
```

```{r}
ggplot(eggs, aes(y = Fat, x = Lab, color = Technician, shape = Sample)) + 
  geom_point(position = position_jitter(width = 0.1, height = 0.0))
```

```{r}
eggs$labtech <- factor(paste0(eggs$Lab, eggs$Technician))
eggs$labtechsamp <- factor(paste0(eggs$Lab, eggs$Technician, eggs$Sample))
```
## INLA

```{r}
formula <- Fat ~ 1 + f(Lab, model = "iid") + f(labtech, model = "iid") + f(labtechsamp, model = "iid")
result <- inla(formula, family = "gaussian", data = eggs)
result <- inla.hyperpar(result)
summary(result)
```
```{r}
# INLA stuff not that relevant
# apar <- 0.5
# bpar <- apar * var(eggs$Fat)
# lgprior <- list(prec = list(prior = "loggamma", param = c(apar, bpar)))
# 
# formula <- Fat ~ 1 + 
#   f(Lab, model = "iid", hyper = lgprior) + 
#   f(labtech, model = "iid", hyper = lgprior) +
#   f(labtechsamp, model = "iid", hyper = lgprior)
# 
# result <- inla(formula, family = "gaussian", data=eggs)
# result <- inla.hyperpar(result)
# summary(result)
```
```{r}
# sigma <- lapply(
#   result$internal.marginals.hyperpar,
#   FUN = function(y) inla.tmarginal(function(x) 1/sqrt(exp(x)), y)
# )
# 
# do.call("rbind", lapply(sigma, FUN = function(x) inla.zmarginal(x, silent = TRUE)))
```
```{r}
# df <- data.frame(do.call("rbind", sigma), "term" = gl(4, 2048, labels = c("epsilon", "Lab", "Tech", "Samp")))
# 
# ggplot(df, aes(x = x, y = y, col = term)) + 
#   geom_line() +
#   labs(x = "Standard deviation", y = "Density") + 
#   xlim(0, 0.25)
```

## Stan

The standard parameterisation (see `nested.stan`):

```{r}
level1 <- as.numeric(eggs$Lab)
level2 <- as.numeric(eggs$labtech)
level3 <- as.numeric(eggs$labtechsamp)

sdscal <- sd(eggs$Fat)

eggdat <- list(
  N = nrow(eggs),
  N1 = max(level1),
  N2 = max(level2),
  N3 = max(level3),
  y = eggs$Fat,
  level1 = level1,
  level2 = level2,
  level3 = level3,
  sdscal = sdscal
)
```

```{r message=FALSE, warning=FALSE}
fit <- rstan::stan(file = "stan/nested.stan", data = eggdat, iter = 2000)
rstan::get_elapsed_time(fit)
```

```{r}
bayesplot::mcmc_trace(fit, pars = c("sigma", "sigma1", "sigma2", "sigma3"))
```

Expect to be observing negative correlation between the precision parameters.
There is only a finite amount of variance to explain so they are in competition with each other.

```{r}
np <- bayesplot::nuts_params(fit)

bayesplot::mcmc_pairs(
  fit, 
  pars = c("sigma", "sigma1", "sigma2", "sigma3"), 
  off_diag_args = list(size = 1, alpha = 0.2),
  np = np
)
```

### Reparametrise

Trying a convolutional reparameterisation analogous to the BYM2 reparameterisation of BYM (see `nested-convolution.stan`):

```{r message=FALSE, warning=FALSE}
fit2 <- rstan::stan(file = "stan/nested-convolution.stan", data = eggdat, iter = 2000)
rstan::get_elapsed_time(fit2)
```

```{r}
bayesplot::mcmc_trace(fit2, pars = c("sigma", "sigmaphi", "pi1", "pi2"))
```
```{r}
np2 <- bayesplot::nuts_params(fit2)

bayesplot::mcmc_pairs(
  fit2, 
  pars = c("sigma", "sigmaphi", "pi1", "pi2"), 
  off_diag_args = list(size = 1, alpha = 0.2),
  np = np2
)
```

Slower but no divergent transitions.

Are they getting the same answer?

```{r}
su_inla <- c(result$summary.random$Lab$mean, result$summary.random$labtech$mean, result$summary.random$labtechsamp$mean)
u_stan <- rstan::summary(fit, pars = c("u1", "u2", "u3"))$summary[, "mean"]
u_stan2 <- rstan::summary(fit2, pars = c("u1", "u2", "u3"))$summary[, "mean"]
```
```{r}
plot(u_stan, u_stan2)
abline(a = 0, b = 1)
```

```{r}
sessionInfo()
```